Variant calling with transformers

### 2022-05-12 manuscript/manuscript.tex
manuscript changes

Variant discovery tools use a variety of statistical techniques to identify variants. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) examined each alignment column individually to determine genotype. Because these tools do not realign reads, they rely on the alignments produced by read mapping tools such as BWA (Li \& Durbin 2009). Later tools, such as the HaplotypeCaller from the GATK and Strelka 2 (Kim et al. 2018) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. The HaplotypeCaller tool constructs de Bruijn graphs from k-mers present in the reads in a region, realigns reads to the graphs, and uses a pair Hidden Markov Model (HMM) to determine the most likely path through the graph.

More recent tools have incorporated elements of deep learning to further improve accuracy. DeepVariant (Poplin et al. 2018) builds on the HaplotypeCaller approach, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections. HELLO (Ramachandran et al. 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input. While these tools deliver accurate results, they rely heavily on hand-crafted statistical algorithms and heuristics to generate and assess candidate alleles.

Here, we show that none of the bespoke statistical apparatus developed in earlier tools is required for accurate variant identification, and that a robust variant detection model can be learned directly from raw NGS data. To achieve this, we recast variant detection as a sequence-to-sequence translation problem. The series of bases aligning to a given reference position is a single input token, and the sequence of these alignment columns constitutes the input to the model. The output of the model is the two predicted haplotype sequences in the sample. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). Transformer models have achieved state-of-the-art results in many sequence-to-sequence problems,

@@

@@

Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region. Any genomic position in which at least two reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant. Positions closer than 100bp are merged into a single region.

@@

@@

>>>>

All variant discovery tools must address two challenges. First, they must identify candidate alleles in a region. Then, given a set of allele candidates, they must assess the likelihood of each candidate (or pairs of candidates in the diploid case), to determine the most likely alleles. Early callers, such as samtools / mpileup (Li et al. 2009) and the UnifiedGenotyper tool from the Genome Analysis ToolKit (GATK, De Pristo et al. 2011) rely on the read aligner to generate candidate allele candidates, and utilize ad-hoc heuristics and thresholds to determine likely alleles. Later tools (e.g. Poplin et al. 2018, Kim et al. 2018, Cooke et al. 2021) incorporated local re-alignment of reads in each region of interest, resulting in a significant improvement in the precision and specificity of variant calls. For instance, the HaplotypeCaller tool identifies candidate haplotypes by constructing de Bruijn graphs from k-mers present in the reads, and assesses haplotype likelihoods with a pair Hidden Markov Model (HMM).

More recent tools have incorporated elements of deep learning into the allele likelihood calculation. DeepVariant (Poplin et al. 2018) builds on the HaplotypeCaller approach, but adds a Convolutional Neural Net (CNN) to classify variants as true or false positive detections. HELLO (Ramachandran et al. 2020), designed to work on hybrid short- and long-read datasets, employs a mixture-of-experts approach with separate 1-dimensional convolutions across the read and allele dimensions of the input.

Here, we explore a new approach to variant detection that replaces the bespoke statistical techniques used in previous tools with a single, end-to-end machine learning model. We recast variant detection as a sequence-to-sequence translation problem in which the input sequence is the series of bases aligning to a single genomic location, and the output sequences are the two predicted haplotypes. The sequence-to-sequence translation is performed by a transformer-based deep neural network (Vaswani et al. 2017). We demonstrate that this approach identifies real genomic variants with sensitivity equivalent to current top-performing callers, albeit with slightly lower specificity.

@@

For all analyses described here, we obtain training data only from the human autosomes 1-20, and hold out chromosomes 21 and 22 for model evaluation.

@@

Given an alignment file in BAM or CRAM format, we first identify regions where a potential variant might exist, and then run forward passes of the model using a sliding window over the region. Any genomic position in which at least three reads contain a base that differs from the reference or an indel are flagged as potentially containing a variant. Positions closer than 100bp are merged into a single region.

@@

\vspace{8pt}
Cooke, Daniel P., David C. Wedge, and Gerton Lunter. "A unified haplotype-based method for accurate and comprehensive variant calling." Nature biotechnology 39.7 (2021): 885-892.

@@

\vspace{8pt}
Poplin, Ryan, et al. "Scaling accurate genetic variant discovery to tens of thousands of samples." BioRxiv (2018): 201178.
